Measles risk assessment and cost interim report\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Nigel P French}

### 2014-07-02 interimreport1.Rnw
Revise interim report Rnw with mapdata
Some of the functions for creating maps were not working through Rnw
Seave, so I made separate results files to work from.

The interim report begins to include material from previous reports and
work on R0.

@@

<<frequentist>>=
library(mcmc)
data(logit)
out <- glm(y ~ x1 + x2 + x3 + x4, data = logit,
    family = binomial())
summary(out)
@

...text

<<log.unnormalized.posterior>>=
x <- logit
x$y <- NULL
x <- as.matrix(x)
x <- cbind(1, x)
dimnames(x) <- NULL

y <- logit$y

lupost <- function(beta, x, y) {
    eta <- x %*% beta
    p <- 1 / (1 + exp(- eta))   # note: works even when eta is Inf or -Inf
    logl <- sum(log(p[y == 1])) + sum(log(1 - p[y == 0]))
    return(logl + sum(dnorm(beta, 0, 2, log = TRUE)))
}
@

@@

<<metropolis-try-1>>=
set.seed(42)    # to get reproducible results
beta.init <- as.numeric(coefficients(out))
out <- metrop(lupost, beta.init, 1e3, x = x, y = y)
names(out)
out$accept
@

...text

\begin{itemize}
\item ...text.
\end{itemize}

...text

\citet{geyer-temp} ...text

<<metropolis-try-2>>=
out <- metrop(out, scale = 0.1, x = x, y = y)
out$accept
out <- metrop(out, scale = 0.3, x = x, y = y)
out$accept
out <- metrop(out, scale = 0.5, x = x, y = y)
out$accept
out <- metrop(out, scale = 0.4, x = x, y = y)
out$accept
@

...text

<<label=metropolis-try-3>>=
out <- metrop(out, nbatch = 1e4, x = x, y = y)
out$accept
out$time
@

Figure~\ref{fig:fig1} (page~\pageref{fig:fig1})
shows the time series plot made by the R statement
<<label=fig1too,include=FALSE>>=
plot(ts(out$batch))
@
\begin{figure}
\begin{center}
<<label=fig1,fig=TRUE,echo=FALSE>>=
<<fig1too>>
@
\end{center}
\caption{Time series plot of MCMC output.}
\label{fig:fig1}
\end{figure}

Another way to look at the output is an autocorrelation plot.
Figure~\ref{fig:fig2} (page~\pageref{fig:fig2})
shows the time series plot made by the R statement
<<label=fig2too,include=FALSE>>=
acf(out$batch)
@
\begin{figure}
\begin{center}
<<label=fig2,fig=TRUE,echo=FALSE>>=
<<fig2too>>
@
\end{center}
\caption{Autocorrelation plot of MCMC output.}
\label{fig:fig2}
\end{figure}

...text
\begin{verbatim}
http://www.stat.umn.edu/~charlie/mcmc/diag.html
\end{verbatim}
says
\begin{quotation}
...text
\end{quotation}
...text
...text
\section{Monte Carlo Estimates and Standard Errors}

<<label=metropolis-try-4>>=
out <- metrop(out, nbatch = 1e2, blen = 100,
    outfun = function(z, ...) c(z, z^2), x = x, y = y)
out$accept
out$time
@

We have added an argument \verb@outfun@ ...text
We need to
use the identity
$$
   \var(X) = E(X^2) - E(X)^2
$$
...text
Hence our \verb@outfun@ returns
\verb@c(z, z^2)@ for an argument (the state vector) \verb@z@.

The \verb@...@ argument to \verb@outfun@ is required, since the
function is also passed the other arguments (here \verb@x@ and \verb@y@)
to \verb@metrop@.

\subsection{Simple Means}

The grand means (means of batch means) are
<<label=metropolis-batch>>=
apply(out$batch, 2, mean)
@
...text
<<label=metropolis-batch-too>>=
foo <- apply(out$batch, 2, mean)
mu <- foo[1:5]
sigmasq <- foo[6:10] - mu^2
mu
sigmasq
@

...text
<<label=metropolis-mcse-mu>>=
mu.mcse <- apply(out$batch[ , 1:5], 2, sd) / sqrt(out$nbatch)
mu.mcse
@
The extra factor \verb@sqrt(out$nbatch)@ arises because the batch means
have variance $\sigma^2 / b$ where $b$ is the batch length, which is
\verb@out$blen@,
whereas the overall means \verb@mu@ have variance $\sigma^2 / n$ where
$n$ is the total number of iterations, which is \verb@out$blen * out$nbatch@.

\subsection{Functions of Means}

To get the MCSE for the posterior variances we apply the delta method.
Let $u_i$ denote the sequence of batch means of the first kind for one
parameter and $\bar{u}$ the grand mean (the estimate of the posterior mean
of that parameter),
let $v_i$ denote the sequence of batch means of the second kind for the
same parameter and $\bar{v}$ the grand mean (the estimate of the posterior
second absolute moment of that parameter), and let $\mu = E(\bar{u})$ and
$\nu = E(\bar{v})$.  Then the delta method linearizes the nonlinear function
$$
   g(\mu, \nu) = \nu - \mu^2
$$
as
$$
   \Delta g(\mu, \nu) = \Delta \nu - 2 \mu \Delta \mu
$$
saying that
$$
   g(\bar{u}, \bar{v}) - g(\mu, \nu)
$$
has the same asymptotic normal distribution as
$$
   (\bar{v} - \nu) - 2 \mu (\bar{u} - \mu)
$$
which, of course, has variance \verb@1 / nbatch@ times that of
$$
   (v_i - \nu) - 2 \mu (u_i - \mu)
$$
and this variance is estimated by
$$
   \frac{1}{n_{\text{batch}}} \sum_{i = 1}^{n_{\text{batch}}}
   \bigl[ (v_i - \bar{v}) - 2 \bar{u} (u_i - \bar{u}) \bigr]^2
$$
So
<<label=metropolis-mcse-sigmasq>>=
u <- out$batch[ , 1:5]
v <- out$batch[ , 6:10]
ubar <- apply(u, 2, mean)
vbar <- apply(v, 2, mean)
deltau <- sweep(u, 2, ubar)
deltav <- sweep(v, 2, vbar)
foo <- sweep(deltau, 2, ubar, "*")
sigmasq.mcse <- sqrt(apply((deltav - 2 * foo)^2, 2, mean) / out$nbatch)
sigmasq.mcse
@
does the MCSE for the posterior variance.

Let's just check that this complicated \verb@sweep@ and \verb@apply@ stuff
does do the right thing.
<<label=metropolis-mcse-sigmasq-too>>=
sqrt(mean(((v[ , 2] - vbar[2]) - 2 * ubar[2] * (u[ , 2] - ubar[2]))^2) /
    out$nbatch)
@

\paragraph{Comment} Through version 0.5 of this vignette it contained
an incorrect procedure for calculating this MCSE, justified by a handwave.
Essentially, it said to use the standard deviation of the batch means called
\verb@v@ here, which appears to be very conservative.

\subsection{Functions of Functions of Means}

If we are also interested in the posterior standard deviation
(a natural question, although not asked on the exam problem),
the delta method gives its standard error in terms of that
for the variance
<<label=metropolis-mcse-sigma>>=
sigma <- sqrt(sigmasq)
sigma.mcse <- sigmasq.mcse / (2 * sigma)
sigma
sigma.mcse
@

\section{A Final Run}

...text...
(the exam problem directed ``use a long enough run of your Markov chain
sampler so that the MCSE are less than 0.01'')
<<label=metropolis-try-5>>=
out <- metrop(out, nbatch = 5e2, blen = 400, x = x, y = y)
out$accept
out$time
<<metropolis-batch-too>>
<<metropolis-mcse-mu>>
<<metropolis-mcse-sigmasq>>
<<metropolis-mcse-sigma>>
@
...text... \verb@xtable@ command in the \verb@xtable@ library.

First the posterior means,
\begin{table}[ht]
\caption{Posterior Means}
\label{tab:mu}
\begin{center}
<<label=tab1,echo=FALSE,results=tex>>=
foo <- rbind(mu, mu.mcse)
dimnames(foo) <- list(c("estimate", "MCSE"),
    c("constant", paste("$x_", 1:4, "$", sep = "")))
library(xtable)
print(xtable(foo, digits = rep(4, 6),
    align = c("l", rep("c", 5))), floating = FALSE,
    caption.placement = "top",
    sanitize.colnames.function = function(x) return(x))
@
\end{center}
\end{table}
then the posterior variances (table on page~\pageref{tab:sigmasq}),
\begin{table}[ht]
\caption{Posterior Variances}
\label{tab:sigmasq}
\begin{center}
<<label=tab1,echo=FALSE,results=tex>>=
foo <- rbind(sigmasq, sigmasq.mcse)
dimnames(foo) <- list(c("estimate", "MCSE"),
    c("constant", paste("$x_", 1:4, "$", sep = "")))
library(xtable)
print(xtable(foo, digits = rep(4, 6),
    align = c("l", rep("c", 5))), floating = FALSE,
    caption.placement = "top",
    sanitize.colnames.function = function(x) return(x))
@
\end{center}
\end{table}
and finally the posterior standard deviations
(table on page~\pageref{tab:sigma}).
\begin{table}[ht]
\caption{Posterior Standard Deviations}
\label{tab:sigma}
\begin{center}
<<label=tab1,echo=FALSE,results=tex>>=
foo <- rbind(sigma, sigma.mcse)
dimnames(foo) <- list(c("estimate", "MCSE"),
    c("constant", paste("$x_", 1:4, "$", sep = "")))
library(xtable)
print(xtable(foo, digits = rep(4, 6),
    align = c("l", rep("c", 5))), floating = FALSE,
    caption.placement = "top",
    sanitize.colnames.function = function(x) return(x))
@
\end{center}
\end{table}

Note for ...
<<label=time,echo=FALSE,results=tex>>=
cat(out$time[1], "\n")
@
...text...

\section{New Variance Estimation Functions}

A new function \texttt{initseq} estimates variances in the Markov chain
central limit theorem (CLT) following the methodology introduced by
\citet[Section~3.3]{practical}.  These methods only apply to scalar-valued
functionals of
reversible Markov chains, but the Markov chains produced by the \texttt{metrop}
function satisfy this condition, even, as we shall see below, when batching
is used.

...text.. AR(1) ...text... \texttt{initseq}.
<<x>>=
n <- 2e4
rho <- 0.99
x <- arima.sim(model = list(ar = rho), n = n)
@
...text...

Define
\begin{equation} \label{eq:little}
   \gamma_k = \cov(X_i, X_{i + k})
\end{equation}
where the covariances refer to the stationary Markov chain having the
same transition probabilities as \texttt{x}.  Then the variance in the CLT
is
$$
   \sigma^2 = \gamma_0 + 2 \sum_{k = 1}^\infty \gamma_k
$$
\citep[Theorem~2.1]{practical}, that is,
$$
   \bar{x}_n \approx \text{Normal}\left(\mu, \frac{\sigma^2}{n}\right),
$$
where $\mu = E(X_i)$ is the quantity being estimated by MCMC (in this
toy problem we know $\mu = 0$).

Naive estimates of $\sigma^2$ obtained by plugging in empirical
estimates of the gammas do not provide consistent estimation
\citep[Section~3.1]{practical}.  Thus the scheme implemented
by the R function \texttt{initseq}.  Define
\begin{equation} \label{eq:big}
   \Gamma_k = \gamma_{2 k} + \gamma_{2 k + 1}
\end{equation}
\citet[Theorem~3.1]{practical} says that $\Gamma_k$ ...text..
\citet[Section~3.3]{practical} ...text.. Theorem~3.2.

Figure~\ref{fig:gamma} (page~\pageref{fig:gamma})
shows the time series plot made by the R statement
<<label=figgamtoo,include=FALSE>>=
out <- initseq(x)
plot(seq(along = out$Gamma.pos) - 1, out$Gamma.pos,
        xlab = "k", ylab = expression(Gamma[k]), type = "l")
lines(seq(along = out$Gamma.dec) - 1, out$Gamma.dec, lty = "dotted")
lines(seq(along = out$Gamma.con) - 1, out$Gamma.con, lty = "dashed")
@
\begin{figure}
\begin{center}
<<label=figgam,fig=TRUE,echo=FALSE>>=
<<figgamtoo>>
@
\end{center}
\caption{Plot ``Big Gamma'' defined by \eqref{eq:little} and \eqref{eq:big}.
Solid line, initial positive sequence estimator.
Dotted line, initial monotone sequence estimator.
Dashed line, initial convex sequence estimator.}
\label{fig:gamma}
\end{figure}
...text... Figure~\ref{fig:gamma}, ...text... $\sigma^2$, .... 
<<assvar>>=
out$var.con
(1 + rho) / (1 - rho) * 1 / (1 - rho^2)
@
where for comparison we have given the exact theoretical value of $\sigma^2$,
...text...
<<batx>>=
blen <- 5
x.batch <- apply(matrix(x, nrow = blen), 2, mean)
bout <- initseq(x.batch)
@
Because the batch length is too short, the variance of the batch means
does not estimate $\sigma^2$.  We must account for the autocorrelation
of the batches, shown in Figure~\ref{fig:gambat}.
<<label=figgambattoo,include=FALSE>>=
plot(seq(along = bout$Gamma.con) - 1, bout$Gamma.con,
        xlab = "k", ylab = expression(Gamma[k]), type = "l")
@
\begin{figure}
\begin{center}
<<label=figgambat,fig=TRUE,echo=FALSE>>=
<<figgambattoo>>
@
\end{center}
\caption{Plot ``Big Gamma'' defined by \eqref{eq:little} and \eqref{eq:big}
for the sequence of batch means (batch length \Sexpr{blen}).
Only initial convex sequence estimator is shown.}
\label{fig:gambat}
\end{figure}
...text the $\sigma^2$ for the original series.
<<compvar>>=
out$var.con
bout$var.con * blen
@
Another way to look at this is that the MCMC estimator of $\mu$ is
either \texttt{mean(x)} or \texttt{mean(x.batch)}.  And the variance
must be divided by the sample size to give standard errors.  So either
<<ci-con>>=
mean(x) + c(-1, 1) * qnorm(0.975) * sqrt(out$var.con / length(x))
mean(x.batch) + c(-1, 1) * qnorm(0.975) * sqrt(bout$var.con / length(x.batch))
@
is an asymptotic 95\% confidence interval for $\mu$.  Just divide by
the relevant sample size.

>>>>

<<echo=FALSE>>=
# haven't had time to sort out which packages I need
library(maps)       # Provides functions that let us plot the maps
library(mapdata)    # Contains the hi-resolution points that mark out the countries.
library(rworldmap)
library(adehabitat)
library(rgdal)
library(raster)
library(maps)
library(maptools)
library(sp)
#library(SDMTools)
library(ggplot2)
require(rgeos)
library(ggmap)
library(mapdata)
#gpclibPermit()
library(mapproj)
library(plyr)
library(grid)
library(gridExtra)
#gpclibPermit()
library(utils)
library(rgdal)
#getwd()
#setwd("~/data")
#unzip("TM_WORLD_BORDERS_SIMPL-0.3.zip")
load("world.ggmap.Rda")
n <- length(unique(world.ggmap$id))
id = unique(world.ggmap$id)
id<-as.data.frame(id)
dpdf<-read.csv("mapdata.csv",header=T)
# output top countries

topimmigration <- dpdf[order(-dpdf$immigration),] 
topimmigration[1:10,c(1,4)]

topincidence <- dpdf[order(-dpdf$incidence),] 
topincidence[1:10,c(1,2)]

topvaccine <- dpdf[order(dpdf$cover),] 
topvaccine[1:10,c(1,3)]

toprisk <- dpdf[order(-dpdf$risk),] 
toprisk[1:10,c(1,5)]

np1<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Immigration") +
  geom_map(aes(fill = immigration), map =world.ggmap) +
  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")

np2<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Incidence (per million)") +
  geom_map(aes(fill = incidence), map =world.ggmap) +
  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")

np3<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Vaccination cover (%)") +
  geom_map(aes(fill = cover), map =world.ggmap) +
  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
  scale_fill_gradient("",low = "red", high = "blue", guide = "colorbar")

#grid.arrange(np1, np2, np3, ncol=3, main="Measles")

np4<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Risk 2012") +
  geom_map(aes(fill = risk), map =world.ggmap) +
  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
#np4

#grid.arrange(np4, np1, np2, np3, ncol=2, main="Measles (2012)")

@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 4>>=
print(grid.arrange(np1, np2, np3, ncol=1, main="Measles (2012)"))
@
\end{center}
\caption{Global risk data: Immigration numbers (top); Measles incidence (per million) (middle); Vaccination coverage (bottom)}
\end{figure}

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 10>>=
print(np4)
@
\end{center}
\caption{Risk map from measles incidence and immigration rates}
\end{figure}

@@

\section{Modeling measles epidemics}

A previously-published model of the dynamics of measles infections
in New Zealand is used to evaluate the present vaccination strategy of
MMR1 at 15 months and MMR2 before 5 years. The results show that
achieving coverage of 90\% or better at both vaccination opportunities is
necessary if future epidemics of measles are to be prevented.

The original mathematical model for the dynamics of measles in New Zealand
was prepared in 1996 (Tobias & Roberts 1998). It successfully predicted the
1997 epidemic, which was curtailed by a mass vaccination campaign (Mansoor
et al 1998; Roberts & Tobias 2000). A subsequent extension of this work in
1998 showed that the then current schedule of MMR1 at 15 months and MMR2
at 11 years was insufficient to prevent further epidemics. The schedule was
changed in 2000 with MMR2 now being administered before 5 years (Anon.
2002a).

A variety of similar models for measles vaccination strategies have been developed
by other authors for various regions (Agur et al 1993; Babad et al 1995;
Edmunds et al 2000; Gay et al 1998; Wallinga et al 2001). These have invariably
been based on sets of nonlinear differential equations, and the conclusions
reached have been similar. The differences in the models have been in the details
of the representation of the infectious period, and in the ways in which
the age and contact structures of the population have been specified. As the
time-course of a measles epidemic is short compared to that of demographic
changes, details of the epidemic process are not relevant when examining population
level changes over many years. In fact, it has been shown (Diekmann
& Heesterbeek 2000) that the number infected in an epidemic depends only
on the basic reproduction number (see below), and the numbers that are initially
susceptible in each sub-population. It is therefore important to have a
realistic representation of the population contact structure. The Royal Society
(UK) report on the 2001 epidemic of Foot and Mouth Disease advocated that
contact structures for infectious diseases should be described before epidemics
occur, so that models may be speedily prepared. The report presented the
New Zealand measles model as an example of the successful use of modelling
in epidemiology (Anon. 2002b).

The quantity that determines whether an epidemic will occur is the basic
reproduction number of the infection, R0. This is defined as the expected
number of secondary infections that would arise from a single primary infection
introduced into a fully susceptible population (Anderson & May 1991;
Diekmann & Heesterbeek 2000). Clearly if R0 > 1 an epidemic will occur following
an introduction of infection. The best estimate we had for measles in
New Zealand was R0 = 12.8, the change in the birth rate (56780 p.a. in 2004,
www.stats.govt.nz) could have reduced this slightly to R0 = 12.5 (see results
section). Recall that R0 is calculated in the absence of control measures. We
are interested in two related quantities:

The basic reproduction number of the infection under vaccination, Rv,
is the expected number of secondary infections that would arise from a
single primary infection introduced into a vaccinated population at equilibrium.
This is a robust indicator of the performance of a vaccination
schedule. If Rv < 1 epidemics are prevented.

The case reproduction number of the infection at time t, Rt, is the
expected number of secondary infections that arise from a single infection
at a particular time. This depends on the number in the population who
are susceptible, either through prior infection or vaccination.

The dynamics of the process may be summarised as follows. Assume that
the population has Rt < 1 for measles. As the number of susceptibles in the
population increases, through children that miss their scheduled immunisations,
then Rt increases. An epidemic occurs when Rt has increased above
one. During the epidemic the number of susceptibles and hence Rt decreases.

In order to assess the efficacy of the vaccination campaign in the absence of
precise information on levels of vaccination coverage, we assumed a range of
plausible coverages at MMR1 and MMR2 and determined the corresponding
value of Rv for each. For selected combinations we also numerically solved
the model over a period of 150 years, not as a prediction of what will happen
over that time period but in order to determine what the expected frequency
of epidemics would be if this level of vaccination were to be in maintained. In
all cases, if Rv < 1 epidemics are prevented.

Method
For this exercise we modelled the New Zealand population in the same way
as in Roberts & Tobias (2000). We approximated the population structure by
assuming that the birth rate was constant, and that deaths before the age of
25 could be neglected. Hence the four age groups active in the epidemic have
constant size. The age groups were taken as 6 - 15 months, 15 months - 5 years,
5 years - 11 years and 11 years - 25 years. Hence MMR1 is applied as children
pass from the first to the second group, and MMR2 from the second to the
third. The same contact matrix was used as in Roberts & Tobias (2000). For
a range of vaccination coverages for MMR1 and MMR2 (see Table 1) the basic
reproduction number of the infection under vaccination, Rv, was calculated.
In addition, for each combination of coverages the differential equation model
was solved numerically until the epidemics settled into a regular pattern, and
the inter-epidemic period was determined.

Calculated values of Rv for the various coverage levels of MMR1 and MMR2
that were considered are presented in Table 1.

Table 1: The basic reproduction number of the infection under vaccination,
Rv, for various coverage levels of MMR1 and MMR2. Superscripts refer to
the figure number where numerical solutions are presented. If Rv < 1 then
epidemics will not occur.

When no vaccination is applied, the system settles down to an epidemic every
two years. This corresponds to an estimated R0 = 12.5, similar to that which
was observed in New Zealand in the 1960s (see Fig.1). The minimum level of
vaccination investigated (MMR1 = 80\%, MMR2 = 60\%) resulted in Rv = 1.72
and a pattern of epidemics settling down to one every 12 years (see Fig.2).
The other combinations of MMR1 and MMR2 coverage levels that resulted
in Rv > 1 all resulted in epidemics recurring at intervals of between 10 and
20 years (Figs. 3-7). All other combinations of MMR1 and MMR2 resulted
in Rv > 1 and the system settling down to a steady uninfected state with no
epidemics, see Table 1 and Fig.8 for an example.

Figures 1 & 2. The epidemic pattern with no vaccination (top) and with
MMR1 = 80\%, MMR2 = 60\% (bottom). The continuous curves show the
cumulative numbers of susceptibles in the population in the age groups 6
months-15 months, 15 months - 5 years, 5 years - 11 years and 11 years -25
years, as determined by the model. The spikes show the numbers infectious
during epidemics (×10).

Figures 3 & 4. The epidemic pattern with with MMR1 = 80\%, MMR2 = 70\%
(top) and with MMR1 = 80\%, MMR2 = 80\% (bottom). The continuous
curves show the cumulative numbers of susceptibles in the population in the
age groups 6 months-15 months, 15 months - 5 years, 5 years - 11 years and
11 years -25 years, as determined by the model. The spikes show the numbers
infectious during epidemics (×10).

Figures 5 & 6. The epidemic pattern with MMR1 = 85\%, MMR2 = 60\%
(top) and with MMR1 = 85\%, MMR2 = 70\% (bottom). The continuous
curves show the cumulative numbers of susceptibles in the population in the
age groups 6 months-15 months, 15 months - 5 years, 5 years - 11 years and
11 years -25 years, as determined by the model. The spikes show the numbers
infectious during epidemics (×10).

Figures 7 & 8. The epidemic pattern with MMR1 = 90\%, MMR2 = 60\%
(top) and with MMR1 = 80\%, MMR2 = 85\% (bottom). The continuous
curves show the cumulative numbers of susceptibles in the population in the
age groups 6 months-15 months, 15 months - 5 years, 5 years - 11 years and
11 years -25 years, as determined by the model. The spikes show the numbers
infectious during epidemics (×10).

The model developed by Roberts & Tobias (2000) supported the change in the
immunisation schedule that took effect in January 2001, at which time MMR2
was changed from delivery at 11 years to delivery before the age of five. These
results were in line with those obtained by other authors, for example: Babad
et al (1995) advocated a two-dose schedule for England and Wales, with the
second vaccination given at age four; and Gay et al (1998) recommended a
second vaccination at either 18 months or five years, to complement the first
vaccination at 12 months in Canada. In addition, Agur et al (1993) found that
vaccinating 85\% of susceptible children aged one to seven years at five-yearly
intervals would prevent epidemics in Israel. These authors all agree that two
vaccinations at no less than five years apart are necessary to prevent measles
epidemics.
A different approach was taken by Wallinga et al (2001). These authors took
existing policies in eight European countries and estimated the coverage rates
required to reduce Rv below one. They found that results depended on the
age at delivery, but no strategy succeeded if coverage rates were below approximately
87\%. Our results presented in Table 1 are similar where coverage is
assumed to be the same at MMR1 and MMR2. The results also show the absolute
necessity of maintaining high coverage rates in order to prevent future
epidemics. It is difficult to estimate the proportion of the school-age population
that have been effectively immunised, but this is continuously being
diluted by children who are not immunised. Wallinga et al (2000) noted that
in Italy only MMR1 was offered at 18 months with no second vaccination opportunity,
and that under most plausible assumptions for contact rates even
100\% coverage would be insufficient to prevent epidemics.
A mathematical model for measles vaccination 11
Our final conclusion must depend on how much faith is placed in the model.
The results are consistent with those of other authors, and from Table 1 it
could be deduced that 85\% coverage at MMR1 and MMR2 could be sufficient
to prevent future measles epidemics. However, a study by Glass et al (2004)
in the Netherlands showed that high overall levels of measles vaccination can
obscure pockets of poor coverage, resulting in localised regions with increased
risk of infection. Our results indicate that overall targets or 90% or more will
need to be achieved to prevent future epidemics in New Zealand.

@@

\section{Acknowledgments}
The authors wish to thank Tomasz Kiedrzynski, Lisa Oakley and Nic Aagaard from the Ministry of Health, Ruth Pirie and colleagues from ESR for help in obtaining the appropriate materials for analyses.
